Semantics for the \texorpdfstring{$ \lambda $}{lambda}-calculus

### 2023-12-14 talks/t04-dutch-cats/outline.tex
Start writing the report in earnest

  The formal definition of an algebraic theory is

@@

@@

@@

>>>>

  The formal definition of an algebraic theory is a sequence of sets $ T_n $, you can think of those as containing the terms in $ n $ free variables, with variables $ x_1, \dots, x_n $ in set $ n $, and a substitution. You can view the substitution as taking a term with $ m $ free variables, and $ m $ terms with $ n $ free variables, and substituting the $ g $'s for the free variables in $ f $.

  I have seen structures like algebraic theories in a lot of literature. They are known by different names: Lawvere theories, cartesian operads, abstract clones or algebraic theories.

  Now we are able to start talking formally about models for the $ \lambda $-theory, by extending the notion of algebraic theory. A $ \lambda $-theory is an algebraic theory, together with abstraction functions, that take a term with $ n + 1 $ free variables and give a term with $ n $ free variables, and application functions that go the other way around.

  A somewhat important example of a $ \lambda $-theory is the pure $ \lambda $-calculus.

  We can also consider models with $ \beta $- and/or $ \eta $-equality, by saying that $ \lambda $ after $ \rho $ or the other way around must be the identity.

  I feel like $ \lambda $-theories are a lot less known than algebraic theories. They were not mentioned in the literature that I saw about algebraic theories, and I have only seen the term in this specific paper.

@@

  Now we will talk a bit about semantics, interpretations or denotations, because we want to interpret the terms in our algebraic theories as functions on some set. More specifically, we want to interpret the terms in $ n $ variables as functions that take $ n $ elements of a set and give an element of the set again.

  For example, in the case of our polynomial ring, we can take as the set the rational numbers, and a polynomial in $ n $ variables then gives an $ n $-ary function.

  The formal definition is \dots

@@

  Now we get to the main theorems.

@@

  The first theorem, Scott's representation theorem, states that (\dots).

### 2023-12-14 talks/t04-dutch-cats/slides.tex
Start writing the report in earnest

% \usepackage{stmaryrd}
\usepackage{tikz-cd}
\usepackage{datetime2}

@@

  \frame{\tableofcontents}

@@

  \frame{\tableofcontents[currentsection]}

@@

  \frame{\tableofcontents[currentsection]}

  \subsection{Scott's representation theorem}
  \begin{frame}
    \frametitle{Scott's representation theorem (1980)}

@@

    For $ \lambda: E(X)_{n + 1} \to E(X)_n $, postcomposing with some morphism $ \overline{abs}: (X \to X) \to X $ gives
    \[ \lambda: E(X)_{n + 1} = (X^{n + 1} \to X) \simeq (X^n \to (X \to X)) \xrightarrow{\overline{abs}\ \circ\ -} (X^n \to X) = E(X)_n. \]

    In the same way, we get $ \rho: E(X)_n \to E(X)_{n + 1} $ from a morphism $ \overline{app}: X \to (X \to X) $.
  \end{frame}

@@

  \subsection{The fundamental theorem of the \texorpdfstring{$\lambda$}{lambda}-calculus}
  \begin{frame}
    \frametitle{``The fundamental theorem of the $ \lambda $-Calculus''}

@@

    There is also a functor from $ \Lambda $-algebras to $ \lambda $-theories.

    This functor again uses the endomorphism theory $ E(X) $ for some object $ X $ to construct the $ \lambda $-theory.

@@

  \subsection{The category of retracts}
  \begin{frame}
    \frametitle{The category of retracts}

@@

  \begin{frame}
    \frametitle{Locally cartesian closed}
    This category is cartesian closed: it has products, and `exponentials'. So its morphisms constitute a simply typed $ \lambda $-calculus: we can do \textit{type theory} with the morphisms.

@@

    For a morphism $ f: A \to B $, we have a pullback functor between slice categories $ f^*: R / B \to R / A $. We have a dependent product and sum along $ f $ if $ f^* $ has a right and a left adjoint.

    If the category is \textit{locally cartesian closed}, we have all dependent products, and so all pullback functors have both adjoints.

@@

    But this is too strong a requirement. Not all pullback functors have both adjoints, but some do. $ R $ is \textit{relatively cartesian closed}.

@@

  \section{My contribution}
  \frame{\tableofcontents[currentsection]}

@@

    \begin{itemize}
      \item Learn the background.
      \item Decode the definitions and theorems.
      \item Find examples.
      \item Formalize.
      \item Mechanize.
    \end{itemize}

>>>>

\usepackage{tikz-cd}
\usepackage{datetime2}
\usepackage{stackengine}

\setlength{\parskip}{.5\baselineskip}
\setbeamertemplate{footline}[frame number]

@@

  \begin{frame}[fragile]
    \begin{columns}
      \begin{column}{.5\textwidth}
        \tableofcontents
      \end{column}
      \begin{column}{.5\textwidth}
        \begin{tikzcd}
          \Lambda-algebras \arrow[d, harpoon, shift left]           &                                                   \\
          \lambda-theories \arrow[d] \arrow[u, harpoon, shift left] & \stackanchor{Endomorphism}{theories} \arrow[l, two heads, hook] \\
          \stackanchor{Relatively\ Cartesian}{Closed\ Categories}   &
        \end{tikzcd}
      \end{column}
    \end{columns}
  \end{frame}

@@

  \begin{frame}[fragile]
    \begin{columns}
      \begin{column}{.5\textwidth}
        \tableofcontents[currentsection]
      \end{column}
      \begin{column}{.5\textwidth}
        \begin{tikzcd}
          \Lambda-algebras \arrow[d, harpoon, shift left]           &                                                   \\
          \lambda-theories \arrow[d] \arrow[u, harpoon, shift left] & \stackanchor{Endomorphism}{theories} \arrow[l, two heads, hook] \\
          \stackanchor{Relatively\ Cartesian}{Closed\ Categories}   &
        \end{tikzcd}
      \end{column}
    \end{columns}
  \end{frame}

@@

  \subsection{Scott's representation theorem}

  \begin{frame}[fragile]
    \begin{columns}
      \begin{column}{.5\textwidth}
        \tableofcontents[currentsubsection]
      \end{column}
      \begin{column}{.5\textwidth}
        \begin{tikzcd}
          \Lambda-algebras \arrow[d, harpoon, shift left]           &                                                   \\
          \lambda-theories \arrow[d] \arrow[u, harpoon, shift left] & \stackanchor{Endomorphism}{theories} \arrow[l, two heads, hook] \\
          \stackanchor{Relatively\ Cartesian}{Closed\ Categories}   &
        \end{tikzcd}
      \end{column}
    \end{columns}
  \end{frame}

@@

    We obtain $ \lambda: E(X)_{n + 1} \to E(X)_n $ as
    \[ \lambda: E(X)_{n + 1} = (X^{n + 1} \to X) \simeq (X^n \to X^X) \xrightarrow{\overline{abs}\ \circ\ -} (X^n \to X) = E(X)_n. \]
    for some morphism $ \overline{abs}: X^X \to X $. In the same way, we get $ \rho: E(X)_n \to E(X)_{n + 1} $ from a morphism $ \overline{app}: X \to X^X $.
  \end{frame}

@@

  \subsection{The fundamental theorem of the \texorpdfstring{$\lambda$}{lambda}-calculus}

  \begin{frame}[fragile]
    \begin{columns}
      \begin{column}{.5\textwidth}
        \tableofcontents[currentsubsection]
      \end{column}
      \begin{column}{.5\textwidth}
        \begin{tikzcd}
          \Lambda-algebras \arrow[d, harpoon, shift left]           &                                                   \\
          \lambda-theories \arrow[d] \arrow[u, harpoon, shift left] & \stackanchor{Endomorphism}{theories} \arrow[l, two heads, hook] \\
          \stackanchor{Relatively\ Cartesian}{Closed\ Categories}   &
        \end{tikzcd}
      \end{column}
    \end{columns}
  \end{frame}
  \begin{frame}
    \frametitle{``The fundamental theorem of the $ \lambda $-Calculus''}

@@

    There is also a functor from $ \Lambda $-algebras to $ \lambda $-theories.
    This functor again uses the endomorphism theory $ E(X) $ for some object $ X $ to construct the $ \lambda $-theory.

@@

  \subsection{The category of retracts}

  \begin{frame}[fragile]
    \begin{columns}
      \begin{column}{.5\textwidth}
        \tableofcontents[currentsubsection]
      \end{column}
      \begin{column}{.5\textwidth}
        \begin{tikzcd}
          \Lambda-algebras \arrow[d, harpoon, shift left]           &                                                   \\
          \lambda-theories \arrow[d] \arrow[u, harpoon, shift left] & \stackanchor{Endomorphism}{theories} \arrow[l, two heads, hook] \\
          \stackanchor{Relatively\ Cartesian}{Closed\ Categories}   &
        \end{tikzcd}
      \end{column}
    \end{columns}
  \end{frame}
  \begin{frame}
    \frametitle{The category of retracts}

@@

  \begin{frame}[fragile]
    \frametitle{Relatively cartesian closed}
    This category is cartesian closed: it has products, and `exponentials'. So its morphisms constitute a simply typed $ \lambda $-calculus: we can do \textit{type theory} with the morphisms.

@@

    \begin{center}
      \begin{tikzcd}
        R / A \arrow[d, dashed] \arrow[r, "f^*" description] & R / B \arrow[d, dashed] \arrow[l, "\sum_f"', bend right] \arrow[l, "\prod_f", bend left] \\
        A & B \arrow[l, "f"]
      \end{tikzcd}
    \end{center}

    \textit{Locally cartesian closed}: all pullback functors have both adjoints.

@@

    In $ R $, not all pullback functors have both adjoints, but some do: \textit{relatively cartesian closed}.

@@

  \section{My contribution}
  \begin{frame}[fragile]
    \begin{columns}
      \begin{column}{.5\textwidth}
        \tableofcontents[currentsection]
      \end{column}
      \begin{column}{.5\textwidth}
        \begin{tikzcd}
          \Lambda-algebras \arrow[d, harpoon, shift left]           &                                                   \\
          \lambda-theories \arrow[d] \arrow[u, harpoon, shift left] & \stackanchor{Endomorphism}{theories} \arrow[l, two heads, hook] \\
          \stackanchor{Relatively\ Cartesian}{Closed\ Categories}   &
        \end{tikzcd}
      \end{column}
    \end{columns}
  \end{frame}

@@

    \begin{itemize}
      \item Learn the background.
      \item Decode the definitions and theorems.
      \item Find examples.
      \item Formalize (on paper).
      \item Mechanize.
    \end{itemize}
